---
title: "LA Crime Report"
author: "Zachary Naumann"
date: "7/16/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)
library(lubridate)

# Create dataset from downloaded .csv file
dl <- tempfile()
download.file("https://dl.dropboxusercontent.com/s/isa5k53bp0tsgii/Crime_Data_from_2020_to_Present.csv?dl=1", dl)
la_crime_stats <- as.data.table(read_csv(dl, col_names = TRUE,
                                         col_types = cols('Crm Cd 3' = col_double(),
                                                          'Crm Cd 4' = col_double())))

la_crime_stats <- la_crime_stats %>%
  select(DR_NO, `DATE OCC`, `TIME OCC`, AREA, `AREA NAME`, `Crm Cd`, `Crm Cd Desc`,
         `Vict Age`, `Vict Sex`, `Vict Descent`, `Premis Cd`, `Premis Desc`)

# Designate which crimes are violent
violent_crimes <- c(110,113,121,122,210,220,230,231,235,236,250,251,434,435,436,
                    437,622,623,624,625,626,627,647,753,755,756,761,860,910,920,
                    928,930,943)

# Convert DATE OCC to date type and pull month/day/dow fields from it
la_crime_stats <- la_crime_stats %>%
  mutate(`DATE OCC` = as.Date(`DATE OCC`, "%m/%d/%y"),
         month = month(`DATE OCC`),
         day = day(`DATE OCC`),
         dow = weekdays(`DATE OCC`))

# Reorder dow field
dow_order <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday",
               "Saturday")
la_crime_stats$dow <- factor(la_crime_stats$dow, level = rev(dow_order))

# Reformat TIME OCC field
la_crime_stats <- la_crime_stats %>%
  mutate(`TIME OCC` = substr(as.POSIXct(sprintf("%04.0f", `TIME OCC`),
                                        format="%H%M"), 12, 16))

# Add violent field to dataset
la_crime_stats <- la_crime_stats %>%
  mutate(violent = case_when(
    `Crm Cd` %in% violent_crimes ~ 1,
    TRUE ~ 0
  ))
la_crime_stats$violent <- as.factor(la_crime_stats$violent)

# Generate train and test sets from la_crime_stats dataset
set.seed(1, sample.kind="Rounding")
index <- createDataPartition(la_crime_stats$violent, times = 1, p = 0.2, list = FALSE)
train_set <- la_crime_stats[-index]
test_set <- la_crime_stats[index]
```

# Introduction
The purpose of this report was to determine if there was an effective way to
predict the occurrence of violent crime based on various factors. The dataset
used was specifically focused on crime in Los Angeles in the year 2020. The
dataset, named *la_crime_stats* and containing 276,584 records, was contained in
a data frame with 16 variables, with each element representing a reported crime.
A preview of the dataset is shown below:
```{r preview, echo=FALSE}
head(la_crime_stats)
```
The purpose of each variable is as follows:  
1. *DR_NO* - unique identifier for the specific crime  
2. *DATE OCC* - date when the crime occurred  
3. *TIME OCC* - time when the crime occurred  
4. *AREA* - numeric code for the police district the crime occurred in  
5. *AREA NAME* - name of the police district the crime occurred in  
6. *Crm Cd* - numeric code for the type of crime  
7. *Crm Cd Desc* - name of the type of crime  
8. *Vict Age* - age of the victim  
9. *Vict Sex* - sex of the victim  
10. *Vict Descent* - ethnicity of the victim  
11. *Premis Cd* - numeric code for the type of location the crime occurred in  
12. *Premis Desc* - name of the type of location the crime occurred in  
13. *month* - month of the year, as a number (1-12)  
14. *day* - day of the month, as a number (1-31)  
15. *dow* - day of the week  
16. *violent* - indicator used to specify whether a crime was violent or not
(non-violent = 0, violent = 1)


There were a number of key steps involved in acquiring the necessary data.
Preceding any data analysis, the dataset was partitioned into a training and
test set. These were named *train_set* and *test_set* and contained 221,267 and
55,317 entries respectively. After partitioning the dataset, individual
variables were tested to see if they had any effect on violent crime. Variables
that looked to have an effect were then modeled to determine if they were valid
predictors for the occurrence of violent crime.

# Methods/Analysis
## Dataset Modifications
Several changes were made to the original *la_crime_stats* dataset to facilitate
easier data analysis. Sixteen variables were excluded because they were either
redundant or referred to information that wasn't relevant to the analysis. The
*DATE OCC* field was modified to exclude irrelevant timestamps, and the
*TIME OCC* field was modified to show the time in a standard HH:MM format. In
addition, the *month*, *day*, *dow*, and *violent* fields were added to
streamline analysis.

## Definition of 'Violent'
For the purpose of this study, crimes were considered violent if they involved
force or threats of force. Crimes were grouped together by *Crm Cd* based on
that criterion. Below is the full list of included crimes for reference:
```{r violent_crimes, echo=FALSE}
la_crime_stats %>% filter(`Crm Cd` %in% violent_crimes) %>% distinct(`Crm Cd Desc`)
```

## Data Analysis
### Overall Crime Frequency
The first thing that was checked in the data analysis process was the frequency
of violent and non-violent crimes over the course of the entire year.
```{r crime_frequency_month, echo=FALSE, fig.align='center'}
# Frequency of each crime type per day of the month for the entire year
train_set %>%
  group_by(violent, month, day) %>%
  summarize(n = n()) %>%
  ggplot(aes(x = day, y = month, fill = n)) +
  geom_tile() +
  ggtitle("Frequency of Crimes Per Day of Month for 2020, Faceted by Crime Type") +
  scale_fill_gradient(low = "white", high = "blue") + 
  facet_wrap(~ violent)
# Noticeable drop in crimes around June, coinciding with COVID lockdowns
```
For both violent and non-violent crimes, there was a distinct decrease in
occurrences starting in June, which coincided with the implementation of
lockdowns due to COVID-19.

The frequency of crimes by hour for every day of the week was also examined.
```{r crime_frequency_day, echo=FALSE, fig.align='center'}
# Frequency of each crime type per hour by day of the week
train_set %>%
  mutate(hour = as.numeric(substr(`TIME OCC`, 1,2))) %>%
  group_by(violent, dow, hour) %>%
  summarize(n = n()) %>%
  ggplot(aes(x = hour, y = dow, fill = n)) +
  geom_tile() +
  ggtitle("Frequency of Crimes Per Hour by Day of Week, Faceted by Crime Type") +
  scale_fill_gradient(low = "white", high = "blue") +
  facet_wrap(~ violent)
# All crimes are significantly less common in the early morning,
# most common in the early evening
```
While there was some variation, every day of the week shared a trend toward
increased criminal activity in the early evening and decreased criminal activity
in the early morning. The type of crime did not have a noticeable effect in
either case.
\newpage

### Crimes by Area
The next section of data checked was the relation between crime occurrence and
area.
```{r crime_by_area, echo=FALSE, fig.align='center'}
# Number of crimes committed in each area
train_set %>%
  filter(violent == 0) %>%
  group_by(DR_NO) %>%
  mutate(n = n()) %>%
  ggplot(aes(x = `AREA NAME`, y = n)) +
  geom_bar(stat = "identity") +
  ggtitle("Number of Non-Violent Crime Occurrence by Area") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_y_continuous(labels = scales::comma)
# Top three areas: 77th Street, Southeast, Central
# Bottom three areas: West LA, Devonshire, Northeast

train_set %>%
  filter(violent == 1) %>%
  group_by(DR_NO) %>%
  mutate(n = n()) %>%
  ggplot(aes(x = `AREA NAME`, y = n)) +
  geom_bar(stat = "identity") +
  ggtitle("Number of Violent Crime Occurrence by Area") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_y_continuous(labels = scales::comma)
# Top three areas: Pacific, West LA, N Hollywood
# Bottom three areas: Foothill, Hollenbeck, Rampart
```
Interestingly, the areas where violent and non-violent crimes were most and
least prevalent were completely different. This highlighted area as a potential
predictor for the crime prediction model.
\newpage

### Crimes by Victimology
Victimology was also analyzed for potential relationships with crime occurrence.
```{r victimology, echo=FALSE, fig.align='center'}
# Average age of victims per crime type
train_set %>%
  group_by(DR_NO) %>%
  filter(`Vict Age` %in% c(seq(0:100))) %>%
  ggplot(aes(x = violent, y = `Vict Age`)) +
  geom_jitter() +
  ggtitle("Average Age of Victims by Crime Type") +
  scale_y_continuous(labels = scales::comma)
# Retirement-aged individuals are less likely to be the victims of violent crimes

train_set %>%
  group_by(DR_NO) %>%
  filter(`Vict Sex` %in% c("M","F"), `Vict Age` %in% c(seq(0:100))) %>%
  ggplot(aes(x = violent, y = `Vict Age`)) +
  geom_jitter() +
  ggtitle("Average Age of Victims by Crime, Faceted by Victim Sex") +
  facet_grid(. ~ `Vict Sex`)
# Very little difference between men and women, men are targeted for violent
# crimes a bit longer than women on average, but don't get targeted for
# non-violent crimes until a few years later than women

# Crimes by ethnicity of victims
# Ethnicity Code: A - Other Asian B - Black C - Chinese D - Cambodian F - Filipino
# G - Guamanian H - Hispanic/Latin/Mexican I - American Indian/Alaskan Native J - Japanese
# K - Korean L - Laotian O - Other P - Pacific Islander S - Samoan U - Hawaiian
# V - Vietnamese W - White X - Unknown Z - Asian Indian
train_set %>%
  group_by(`DR_NO`) %>%
  mutate(n = n()) %>%
  ggplot(aes(x = violent, y = n, color = `Vict Descent`)) +
  geom_bar(stat = "identity") +
  ggtitle("Number of Crimes Committed per Crime Type by Victim Ethnicity")
# Black and Hispanic/Latino individuals are equally affected by both violent and
# and non-violent crimes, but other ethnicities are more likely to be targeted
# by non-violent crimes
```
Neither victim age nor sex played a significant role in whether they were
targeted for violent crime. Victim ethnicity played a role when the victims were
neither Black nor Hispanic/Latino.
\newpage

### Crime by Location
The last potential relationship was between the crime and location type.
```{r location, echo=FALSE, fig.align='center'}
# Types of locations where crimes occur per area
train_set %>%
  group_by(`DR_NO`) %>%
  mutate(n = n()) %>%
  ggplot(aes(x = `Premis Cd`, y = n, color = `violent`)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90))
# A vast majority of crimes occur on streets, in single-family dwellings, and
# in multi-unit dwellings (apartments, duplexes, etc.)

train_set %>%
  group_by(`Premis Desc`) %>%
  summarize(n = n()) %>%
  top_n(5) %>%
  pull(`Premis Desc`)
```
A vast majority of both violent and non-violent crimes occurred in 5 types of
locations.

## Predictors and Model
Out of all of the variables analyzed, only the *AREA* field had a significant
relationship with crime occurrence by crime type. Therefore, the model utilized
that field in predicting the occurrence of violent crime. A basic GLM model was
implemented that used the *AREA* field to predict whether a crime was violent
or not.

# Results
```{r glm_model, echo=FALSE}
set.seed(1, sample.kind="Rounding")
fit_glm <- train(violent ~ AREA, train_set, method = "glm")
pred_glm <- predict(fit_glm, test_set)
acc <- mean(pred_glm == test_set$violent)
```
The GLM model predicted whether a crime was violent or not with an accuracy of
`r acc`. This is fairly poor accuracy, being correct only about 2/3 of the time.
Adding other fields to the model either did nothing or made the accuracy worse.
A large part of the problem was that while there were many minor relationships
between various fields and the violent crime rate, nothing fit within a
normal distribution that could be easily modeled.

# Conclusion
The results of this report indicate that the model used was not that effective
at predicting whether a crime was violent or not. There was only one field in
the dataset that was useful in building the model, which hindered its
effectiveness. There were several limitations while working on this model. The
first was the effect COVID-19 had on the data. The sudden, sharp decline in
crime that occurred midway through the year significantly reduced the number of
potential entries that could have led to a more robust analysis. The second
limitation was the fact that only a year of data was used. The source of the
data actually had data going back up to a decade, but the size of the dataset
would have caused problems when trying to run a model. That said, if there was
ever any future work done on this topic, the dataset would probably be expanded
to include several years' worth of data. Also, the model would be changed to
something involving time series analysis, possibly forecasting the rate of
violent crime using historical data.

## Citatations
P, Sumaia. “Los Angeles Crime Data 2010-2020.” Kaggle, 22 June 2021, www.kaggle.com/sumaiaparveenshupti/los-angeles-crime-data-20102020.